---
title: "Introduction to educabr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to educabr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  out.width  = "100%"
)
```

```{r setup, message = FALSE}
library(educabr)
```

## About the package

**educabr** provides, under a single canonical tidy schema, the most
extensive set of harmonised historical series on Brazilian education
available in analytic format:

* **Tertiary (higher-education) enrollment** — 1907 to 2024, **118
  years** of coverage, drawing on seven distinct primary sources
  (IBGE *Statistics of the 20th Century*, Durham, Maduro Junior,
  Kang, INEP Synopsis, INEP Microdata, and the INEP CENSUP Power BI
  panel).
* **Compulsory (fundamental) and upper-secondary enrollment** — 1933
  to 2010, with **race/colour** disaggregation from 1960 onwards
  (Kang, Paese & Felix 2021).
* **Mean years of schooling of the adult population** — 1925 to
  2015, with breakdowns by **sex**, **race/colour**, **macroregion**
  and **state** (Walter & Kang 2024).

Every transformation is auditable: each output row carries a
canonical `source` key (catalogued in
`inst/dict/vocabularies/sources.yaml`) and a `source_note` with the
exact table or chapter of origin.

## Installation

```r
# From GitHub
remotes::install_github("mancano-tales/educabr")
```

## A two-function API

The public interface is intentionally minimal:

```{r}
# Schemas of the two main return shapes
str(get_enrollment(level = "fundamental", year = 1950))
str(get_schooling(year = 1950))
```

Both return long-format tibbles that follow the same canonical
schema (`inst/dict/schema.yaml`), with columns for the geographic
unit, the inequality dimension (race, sex, …), the source, and the
value.

## Case 1 — gross enrollment rate by race/colour

Using `dimension = "race"` we can trace the trajectory of gross
fundamental-education enrollment for the five IBGE race/colour
categories between 1960 and 2010:

```{r}
fund_race <- get_enrollment(
  level     = "fundamental",
  indicator = "rate",
  geo_level = "BR",
  dimension = "race",
  year      = c(1960, 2010)
)

head(fund_race)
```

The gap between the three main categories (`white`, `black`,
`brown`) traces one of the central narratives in the sociological
literature on Brazilian education — relative convergence at
compulsory levels alongside persistent inequality at higher levels.

## Case 2 — multi-source comparison for tertiary enrollment

Brazilian tertiary enrollment has been reconstructed by **multiple
authors** with slightly different methodologies, especially for the
long 1933–2000 period when official sources are sparse. The
**educabr** package keeps **all competing estimates** side by side,
enabling direct comparison:

```{r}
ter_1980 <- get_enrollment(
  level     = "superior",
  year      = 1980,
  network   = "total",
  modality  = "total",
  indicator = "count"
)

ter_1980[, c("source", "value", "source_note")]
```

For 1980 several sources converge on the same value (~1.38 million),
which suggests a stabilised figure in the literature. For other
years estimates diverge — passing `source = "..."` to
`get_enrollment()` is the recommended way to pin down a single
series for secondary analysis.

## Case 3 — the reconstructed totals problem (2000-2008)

Between 2000 and 2008 INEP began collecting distance-learning (EAD)
enrollment in a separate CENSUP table (`tabela7.x`), but **did not
add** it to the total of in-person enrollment published in
`tabela5.x`. This means that headline "total" series for that
interval published by Kang, Durham, Maduro Junior, and INEP's own
synopsis are **systematically undercounted** — by up to ~700,000
enrollments in 2008 by our reckoning.

From 2009 onwards the CENSUP microdata already aggregate in-person
and EAD in the same registry, and the problem resolves itself.

For this transition interval, **educabr** ships **reconstructed
totals** that combine the in-person component of each source with
INEP's published EAD figures. These rows carry `is_derived = TRUE`
and are excluded from the default `get_enrollment()` output; pass
`include_derived = TRUE` to inspect them:

```{r}
recon <- get_enrollment(
  level           = "superior",
  year            = c(2000, 2008),
  network         = "total",
  modality        = "total",
  indicator       = "count",
  include_derived = TRUE
)

recon[recon$is_derived, c("year", "source", "value")]
```

The composite `source` key (e.g.
`kang_paese_felix_2021+inep_sinopse_censup`) makes it explicit that
this is a combination: the **in-person component** comes from the
first source and the **EAD component** comes from the second. The
`source_note` records the exact composition for citation purposes.

> Suggested citation convention: "Reconstructed total computed from
> {in-person source} (in-person) and INEP CENSUP Synopsis (EAD),
> following the methodology documented in `educabr` (Mançano 2026)."

## Case 4 — mean years of schooling by sex

```{r}
schooling_sex <- get_schooling(
  geo_level = "BR",
  dimension = "sex"
)

tail(schooling_sex)
```

This series captures the historical reversal of the *gender gap*: in
1925 men averaged more years of schooling than women; over the 20th
century that advantage inverts, and by 2015 women surpass men in
average attainment.

## Interactive dashboard

The package ships a Shiny dashboard for the three series:

```r
educabr::run_dashboard()
```

The dashboard replicates the multi-source tertiary comparison
interactively. Each tab has a **"View R code"** button that emits
the `educabr` + `ggplot2` snippet needed to reproduce the
visualisation in your local R session — a direct bridge between
dashboard exploration and reproducible scripted analysis.

## Sources and citation

Every series carries a canonical source key (`source`) with the
full reference catalogued in
`inst/dict/vocabularies/sources.yaml`. The seven primary tertiary
sources are:

* Kang, T. H., Paese, L. H. Z., & Felix, N. F. A. (2021). Late and
  unequal. *Revista de Historia Económica* 39(2), 191-218.
  [doi:10.1017/S0212610921000112](https://doi.org/10.1017/S0212610921000112)
* Kang, T. H., & Menetrier, I. (2024). Políticas elitistas e
  despesas públicas em educação. *Estudos Econômicos* 54(3).
  [doi:10.1590/1980-53575434tkim](https://doi.org/10.1590/1980-53575434tkim)
* Kang, T. H., Menetrier, I., & Comim, F. (2024). The side effects
  of a big push growth strategy. *RHE* 42(3), 387-414.
  [doi:10.1017/S0212610924000120](https://doi.org/10.1017/S0212610924000120)
* Walter, J. R., & Kang, T. H. (2024). A new dataset of average
  years of schooling in Brazil. *Economic History of Developing
  Regions* 39(3), 307-336.
  [doi:10.1080/20780389.2024.2417268](https://doi.org/10.1080/20780389.2024.2417268)
* Durham, E. R. (2005). Educação superior, pública e privada. In
  Schwartzman (ed.), *Os desafios da educação no Brasil*, pp. 191-233.
* Maduro Junior, P. R. R. M. (2007). *Taxas de matrícula e gastos em
  educação no Brasil* [MSc dissertation, FGV/EPGE].
  [hdl:10438/110](https://hdl.handle.net/10438/110)
* IBGE (2007). *Estatísticas do Século XX*.
  [seculoxx.ibge.gov.br](https://seculoxx.ibge.gov.br/)
* INEP/MEC. *Sinopse Estatística da Educação Superior* (1995-2008).
* INEP/MEC. *Microdados do Censo da Educação Superior* (2009-2024).
* INEP/MEC. *CENSUP Power BI panel* (2010-2024).

To cite the package itself:

> Mançano, T. (2026). *educabr: Harmonized Historical Series on
> Brazilian Education* (version 0.0.0.9000).
> <https://github.com/mancano-tales/educabr>
